### 2021-01-24 paper/abstract.tex
Trimmed abstract

% Program logics and semantics tell us that in order to derive
% \begin{math}
%   ((S_1\SEMI S_2), \sigma_0) \Downarrow \sigma_2,
% \end{math}
% we derive 
% \begin{math}
%   (S_1, \sigma_0) \Downarrow \sigma_1
% \end{math}
% and then
% \begin{math}
%   (S_2, \sigma_1) \Downarrow \sigma_2.
% \end{math}
Program logics and semantics tell us that when executing ((S1; S2), state0),
we execute (S1, state0) to arrive at state1, then execute (S2, state1) to
arrive at the final state2.

This is a delightfully simple story that can be explained to children.  It is
also a lie.

Processors execute instructions out of order, due to pipelines and caches.
Compilers reorder programs even more dramatically.  All of this reordering is
meant to be unobservable in single-threaded code.  In multi-threaded code,
however, all bets are off.  A formal attempt to understand the resulting mess
is known as a ``relaxed memory model.''

Most of the relaxed memory models that have been proposed are designed to
help us understand whole program execution: they have no compositionality
properties whatsoever.  Recently, denotation models have appeared that treat
\emph{concurrent} execution compositionally.  One such model is ``Pomsets with
Preconditions''.  It remains an open question, however, whether it is
possible to treat \emph{sequential} execution compositionally in such a model,
without overly restricting processors and compilers.

We propose adding families of predicate transformers to Pomsets with
Preconditions.  The resulting model is denotational, supporting both parallel
and sequential composition.  When composing (S1;S2), the predicate
transformer used to validate the precondition of an event in S2 is chosen
based on the dependency order from S1 into this event.  As usual in work on
relaxed memory, we have not handled loops or recursion.

Happily, most of the results expected of a relaxed memory model can be
established by appeal to prior work.  So here we are able to concentrate on
the model itself.  The model is formalized in Agda, where we have established
associativity for sequential composition.

For the memory-model specialist, we retain the good properties of the prior
work on Pomsets with Preconditions, fixing some errors along the way.  These
properties include efficient implementation on ARMv8, support for compiler
optimizations, support for logics that prove the absence of thin-air
behaviors, and a local data race freedom theorem.

>>>>

This paper presents the first semantics for relaxed memory with a
compositional definition of sequential composition.  Previous
definitions of relaxed memory have given detailed treatments of
parallel composition, but have given sequential composition less
attention, often relegating it to a (sometimes speculative)
operational semantics of single-threaded programs. In this paper we show how
sequential composition can be restored to a first-class citizen, by giving
it a denotational semantics in a model of pomsets with preconditions,
extended with a family of predicate transformers. Previous work has shown that
pomsets with preconditions are a model of concurrent composition,
and that predicate transformers are a model of sequential composition.
This is the first paper to show how they can be combined.

% Program logics and semantics tell us that in order to derive
% \begin{math}
%   ((S_1\SEMI S_2), \sigma_0) \Downarrow \sigma_2,
% \end{math}
% we derive 
% \begin{math}
%   (S_1, \sigma_0) \Downarrow \sigma_1
% \end{math}
% and then
% \begin{math}
%   (S_2, \sigma_1) \Downarrow \sigma_2.
% \end{math}
%% Program logics and semantics tell us that when executing ((S1; S2), state0),
%% we execute (S1, state0) to arrive at state1, then execute (S2, state1) to
%% arrive at the final state2.

%% This is a delightfully simple story that can be explained to children.  It is
%% also a lie.

%% Processors execute instructions out of order, due to pipelines and caches.
%% Compilers reorder programs even more dramatically.  All of this reordering is
%% meant to be unobservable in single-threaded code.  In multi-threaded code,
%% however, all bets are off.  A formal attempt to understand the resulting mess
%% is known as a ``relaxed memory model.''

%% Most of the relaxed memory models that have been proposed are designed to
%% help us understand whole program execution: they have no compositionality
%% properties whatsoever.  Recently, denotation models have appeared that treat
%% \emph{concurrent} execution compositionally.  One such model is ``Pomsets with
%% Preconditions''.  It remains an open question, however, whether it is
%% possible to treat \emph{sequential} execution compositionally in such a model,
%% without overly restricting processors and compilers.

%% We propose adding families of predicate transformers to Pomsets with
%% Preconditions.  The resulting model is denotational, supporting both parallel
%% and sequential composition.  When composing (S1;S2), the predicate
%% transformer used to validate the precondition of an event in S2 is chosen
%% based on the dependency order from S1 into this event.  As usual in work on
%% relaxed memory, we have not handled loops or recursion.

%% Happily, most of the results expected of a relaxed memory model can be
%% established by appeal to prior work.  So here we are able to concentrate on
%% the model itself.  The model is formalized in Agda, where we have established
%% associativity for sequential composition.

%% For the memory-model specialist, we retain the good properties of the prior
%% work on Pomsets with Preconditions, fixing some errors along the way.  These
%% properties include efficient implementation on ARMv8, support for compiler
%% optimizations, support for logics that prove the absence of thin-air
%% behaviors, and a local data race freedom theorem.
